SUPPLEMENTARY:\\Correcting for bias in high-throughput sequencing data

### 2011-09-29 response/response.tex

\documentclass{article}
\usepackage{amsmath}
\usepackage{dcj}

\begin{document}

We appreciate the time and effort extended by the anonymous reviewers and the
editorial staff to evaluate the original version of this manuscript (titled
``Correcting for bias in high-throughput sequencing data''), which was submitted
on July 8. Upon review, the second of the two reviewers raised some concerns
leading to its rejection. However, we believe that the bulk of these concerns
are either incongruous with the content of the manuscript or addressable with
very little effort, and therefore do not warrant its rejection.  Here we have
provided a detailed response to these comments, and have included additional
material in an effort to clarify these points.

@@

\begin{quote}
(1) Page 3, 2.2.1 mentions a correction for bias introduced by PCR amplification.
RNA sequencing methods are available where PCR amplification is not used.
Correcting for bias in these cases removes true duplication.  Could the authors
please comment on this?
\end{quote}

First, we emphasize that PCR bias is only one of a number of technical steps
involved in sequencing that potentially introduce bias.  Our method is agnostic
to the source of bias, and because model complexity is penalized, a very sparse,
or even empty model will be trained on a data set without bias.  Thus, if PCR
amplification is responsible for a majority of the bias, and an RNA-Seq method
without PCR amplification is used producing largely bias-free data, then using
the method to correct for bias will have little if any effect.

As a concrete example, we have included new data examining the FRT-Seq protocol
of Mamanova, et al., in which no PCR-amplification is performed during library
preparation. We observe that bias is still present in this data set, albeit at
lower levels than other uncorrected RNAseq data examined.  Our method accounts
for this reduced bias by constructing a much sparser model, relative to the
other RNA-Seq data sets we have examined.

We have included additional material in Supplementary Section 10 exploring this
point in detail. We first prove an upper bound on the probability of the model
predicting the presence of bias when there is none, demonstrating that is less
than 0.0004 when more than 10,000 reads are used. We then investigate this
empirically by simulating random reads, drawn uniformly from exonic sequences
within the human genome.  We trained our model 100 times with 10,000 simulated
reads, and in no instance was bias predicted (i.e., a non-empty model trained).
We repeated this with using only 100 reads, and in only 3 out of 100 trials was
bias incorrectly predicted.

This is indicative that the default BIC scoring, as predicted by statistical
theory, is adequately protecting against false discovery of bias when there is
none, and there is very little risk in applying the method to an unbiased
experiment.

@@

We regret that we were not more explicit on this point. At least
600,000 reads on each data set, but as many as 23 million were used:
see the revised table X. (As noted in the paper, we did reimplement
the Genominator method in order to do this.)

We do not wish to ``pick on'' Hansen, et al.  Theirs was a pioneering
paper and presents a simple, intuitive approach for tackling the
problem they identified.  Nevertheless, their brute-force method
performed poorly on every data set we tested, even the largest ones,
suggesting that simply throwing more data at a suboptimal solution
will not adequately solve the problem.  In our opinion, the results we
presented were unambiguous: the other methods perform far better with
far fewer reads.

%Moreover, even if the brute-force method were effective if trained
%with many hundreds of millions of reads, many (possibly most) existing
%data sets are not this large, rendering it generally impractical.

Understanding this failure of a simple, appealing method is important
for making progress, and we feel that our paper provides a theoretical
basis for understanding these differences in efficacy, and we have
amplified these points in the revision.  Intuitively, one expects
model accuracy to increase as the number $n$ of training reads
increases, but to decrease as the number $r$ of model parameters
grows, since the training data is ``spread more thinly.''  For the
particular case of estimating a multinomial distribution (as in
Hansen), Supplementary Section YY quantifies this intuition very
precisely, showing that the expected KL divergence of the learned
model with respect to the true distribution converges to $(r-1)/(2n)$,
where $r=4^7=16384$ is the number of model parameters learned.  The
same analysis does not directly apply to the MART, GLM or BN models,
but we expect their behavior to be qualitatively similar.  Thus, to
the extent that a more parsimonious model such as ours adequately
captures the structure of the data, it is potentially orders of
magnitude more accurate for a given quantity of training data, simply
by virtue of having only a few hundred parameters learned from counts
(Figure 3), as opposed to tens of thousands.

%[KL divergence between
%two distributions is inversely proportional to the number of samples
%one needs to confidently test the hypothesis that the samples arise
%from one distribution versus the other; see Supplementary Section XX.]
%% TODO: why the brackets here?

@@

Instead, we presented data (in Section 3.2) that answers the
reviewer's question (i.e., ``does this unevenness go away with the
proposed adjustment?'') in rigorous and comprehensive manner. We
implemented a Poisson regression metric, which quantitatively measures
``uniformity'' by calculating the likelihood that the reads sampled
across a given exon were sampled from a uniform distribution.  Using
this, we performed a large cross-validation study proving the
effectiveness of our method at increasing uniformity across thousands
of exons drawn from a diverse group of data sets, rather than
qualitatively across one exon.  In broad terms, our method increases
likelihood of the uniform model by XXX-fold across thousands of exons
(Figure 6).

In this revision, as suggested, we have revised Figure 2 to show an
example of this increase in uniformity.  We believe the figure is a
nice, graphic illustration of the benefits of our method and thank the
reviewer for causing us to reconsider it.  We stress, however, that
this figure represents a minute subset of the analysis that had
previously been presented, and is included only as an illustration.
The results in sections 3.1 and especially in 3.2 are far more
comprehensive.

\begin{quote}
- Also, as is evident from that figure, for some of the experimental
protocols, some regions within highly expressed exons get vanishingly
small read coverage.  These are the really hard situations, and much
more worrying than instances where coverage just fluctuates a bit.
How does the proposed scaling adjustment of Section 2.1 work in these
places (i.e. when $s_{i}$ is very small)?
\end{quote}

In Figure 2, we now indicate that the adjustments made by the method to read
counts tend to be small. In our test, 95\% of a sample of 10 million predicted
bias coefficients were between 0.07 and 4.27. The lack of large adjustments is
also reflected indirectly in the results in Section 3.  This is reflected
indirectly in the results of Section 3: extremely large adjustments would result
in large variance of the predictions, leading to poor performance in our
evaluations, which we do not see.

@@

\begin{quote}
2. The chosen Bayesian network is rather complicated, and a spectrum
of alternative methods exists between that and the brute force
approach of Hansen (say, lasso or svm regression). In absence of a
more meaningful benchmark, the choice of method seems arbitrary.
\end{quote}

Our motivation for adopting a Bayesian network model is the very
natural probabilistic interpretation that we have described in Section
2.1. We disagree with the reviewer's opinion that Bayesian networks
are inherently more complicated than Lasso regression or support
vector machines.  In this application, the latter two methods would
generate output that is arguably less transparent and interpretable,
and would neither be significantly simpler to implement, nor involve
fewer parameters.  Our hypotheses (above) concerning the reasons for
the weakness of Hansen's method illustrate the benefit of this
approach; we do not believe svm regression, say, would have yielded
similar insights.

Naturally, it is possible that another model might perform as well or better
than the one presented here, yet it is unrealistic or impossible to prove that
this is not case. What we have demonstrated is a model that is well reasoned,
efficiently implemented, and effective at reducing sequence bias in RNA-Seq.

\subsection*{Other Changes}

We have modified the title to more clearly acknowledge the existence of
the prior work upon which we build.

### 2011-09-29 seqbias-supplement.tex
(almost) final draft

%% TODO: coptions/labels below this point

@@

Measuring nucleotide frequencies, we find the bias to be significantly less
than that observed the RNA-Seq data sets we tested, but the data was by no means
unbiased.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig/freqs_cao.pdf}
\end{center}
% TODO: caption
\label{fig:trimmedfreqs}
\end{figure}

@@

Below, we plot directly the positional KL divergence, computed in tho same
manner described in the Results section of the paper.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.5\textwidth]{fig/cao-kl.pdf}
\end{center}
% TODO: caption
\label{fig:trimmedfreqs}
\end{figure}

We see that our method is effective at reducing the bias in this ChIP-Seq data.

@@

To evaluate the extent to which the observed bias is caused by PCR amplification
in the library preparation stage, we analyzed data from the FRT-Seq protocol
developed by Mamanova, et. al. \cite{Mamanova2010}. FRT-Seq avoids the
amplification step during library preparation with reverse transcription
occuring on the flowcell surface. We obtained data generated by Mamanova, et.
al., from the European Nucleotide Archive with accession code ERR007689, and
mapped them to the hg19 assembly of the human genome with Bowtie
\cite{Langmead2009}. The nucleotide frequencies and divergence are plotted
below.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.6\textwidth]{fig/frt-all.pdf}
\end{center}
% TODO: caption
\label{fig:trimmedfreqs}
\end{figure}

It is clear from this analysis that the FRT-Seq method, as implemented to
generate this data, is not without bias, yet there is significant divergence at
only a small number of positions surrounding the read start. When our method is
trained on this data we obtain the suitably sparse model pictured below.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.4\textwidth]{fig/frt-graph.pdf}
\end{center}
% TODO: caption
\label{fig:trimmedfreqs}
\end{figure}

@@

Measuring differential expression or isoform switching is a primary application of
RNA-Seq. If the bias were inconsistent between replicates, it would call into
question the accuracy of such tests. In our experiments, we have observed the
bias to be mostly, but not entirely consistent between replicates. Below, we
plot the frequencies from the four runs in the Trapnell data set.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig/replicates.pdf}
\end{center}
% TODO: caption
\label{fig:trimmedfreqs}
\end{figure}

@@

\subsection{TODO: Title}

@@

\section{TODO: title}

@@

As discussed in Supplementary Section 9.2, $2n D$ is approximated by a $chi^2$
distribution with 3 degrees of freedom for large $n$, and so we can estimate
this probability for various values of $n$. This is shown in Figure
\ref{fig:nepr}. From this figure we can see, for example that training with more
than 10,000 reads results in an incorrect non-empty model with probability less
than 0.0004.

In the low end, the upper bound on non-empty model probability is much higher.
In our tests, this upper bound is not tight for a small number of reads. We
trained the model on 100 sets of 100 simulated reads, drawn randomly at uniform
from exonic sequences and in only 3 trials was a non-empty model trained, far
less than the upper bound prediction of 23 out of 100. This test was repeated
using sets of 50,000 reads, and in no trial was a non-empty model trained.

@@

\begin{figure}[H]
\centerline{
\includegraphics[width=0.8\textwidth]{fig/nepr.pdf}}
\caption{Plotted is the upper bound on the probability of training a non-empty
model when there is no real bias in the experiment when the model is trained
with various number of reads $n$. We fixed the number of nucleotides being
considered as $h = 41$, as this what was used for the evaluation in Section 3.}
\ref{fig:nepr}
\end{figure}

>>>>

@@

Measuring nucleotide frequencies, we find the bias to be significantly less
than that observed the RNA-Seq data sets we tested, but the data was by no means
unbiased, as shown in Figure \ref{fig:freqscao}.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig/freqs_cao.pdf}
\end{center}
\caption{Nucleotide frequencies and KL divergence for the Cao data set.}
\label{fig:freqscao}
\end{figure}

@@

Figure \ref{fig:caokl}, we plot directly the positional KL divergence, computed
in tho same manner described in the Results section of the paper.  We see that
our method is effective at reducing the bias in this ChIP-Seq data.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.5\textwidth]{fig/cao-kl.pdf}
\end{center}
\caption{KL divergence for the Cao data set, after adjusting read counts for
bias.}
\label{fig:caokl}
\end{figure}

@@

To evaluate the extent to which the observed bias is caused by PCR amplification
in the library preparation stage, we analyzed data from the FRT-Seq protocol
developed by Mamanova, et. al. \cite{Mamanova2010}. FRT-Seq avoids the
amplification step during library preparation with reverse transcription
occuring on the flowcell surface. We obtained data generated by Mamanova, et.
al., from the European Nucleotide Archive with accession code ERR007689, and
mapped them to the hg19 assembly of the human genome with Bowtie
\cite{Langmead2009}. The nucleotide frequencies and divergence are plotted
in Figure \ref{fig:frtall}.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.6\textwidth]{fig/frt-all.pdf}
\end{center}
\caption{Nucleotide frequencies and KL divergence of the Mamanova data set.}
\label{fig:frtall}
\end{figure}

It is clear from this analysis that the FRT-Seq method, as implemented to
generate this data, is not without bias, yet there is significant divergence at
only a small number of positions surrounding the read start. When our method is
trained on this data we obtain the suitably sparse model pictured in Figure
\ref{fig:frtgraph}.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.4\textwidth]{fig/frt-graph.pdf}
\end{center}
\caption{The graphical model learned by training our method on the Mamanova data
set.}
\label{fig:frtgraph}
\end{figure}

@@

Measuring differential expression or isoform switching is a primary application of
RNA-Seq. If the bias were inconsistent between replicates, it would call into
question the accuracy of such tests. In our experiments, we have observed the
bias to be mostly, but not entirely consistent between replicates. In Figure
\ref{fig:replicates} we
plot the frequencies from the four runs in the Trapnell data set.

\begin{figure}[H]
\begin{center}
\includegraphics[width=0.7\textwidth]{fig/replicates.pdf}
\end{center}
\caption{Nucleotide frequencies of replicates in the Trapnell data set.}
\label{fig:replicates}
\end{figure}

@@

\subsection{Overview}

@@

\section{False Discovery of Bias in Unbiased Experiments}

@@

As discussed in Supplementary Section 9.2, $2n D$ is approximated by a $\chi^2$
distribution with 3 degrees of freedom for large $n$, and so we can estimate
this probability for various values of $n$. This is shown in Figure
\ref{fig:nepr}. From this figure we can see, for example that training with more
than 10,000 reads results in an incorrect non-empty model with probability less
than 0.0004.

In the low end, the upper bound on non-empty model probability is much higher.
In our tests, this upper bound is not tight for a small number of reads. We
trained the model on 100 sets of 100 simulated reads, drawn randomly at uniform
from exonic sequences and in only 3 trials was a non-empty model trained, far
less than the upper bound prediction of 23 out of 100. This test was repeated
using sets of 10,000 reads, and in no trial was a non-empty model trained.

@@

\begin{figure}[H]
\centerline{
\includegraphics[width=0.8\textwidth]{fig/nepr.pdf}}
\caption{Plotted is the upper bound on the probability of training a non-empty
model when there is no real bias in the experiment when the model is trained
with various number of reads $n$. We fixed the number of nucleotides being
considered as $h = 41$, as this what was used for the evaluation in Section 3.}
\label{fig:nepr}
\end{figure}
